---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# svi

<!-- badges: start -->
[![R-CMD-check](https://github.com/YimingMiao/svi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/YimingMiao/svi/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/YimingMiao/svi/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/YimingMiao/bis620.2022/actions/workflows/test-coverage.yaml)
[![Codecov test coverage](https://codecov.io/gh/YimingMiao/svi/branch/main/graph/badge.svg)](https://app.codecov.io/gh/YimingMiao/svi?branch=main)
[![lint](https://github.com/YimingMiao/svi/actions/workflows/lint.yaml/badge.svg)](https://github.com/YimingMiao/svi/actions/workflows/lint.yaml)
<!-- badges: end -->

## Installation

You can install the development version of svi from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("YimingMiao/svi")
```
## Description

`svi` provides tools to study the relationship between social vulnerability index and disease mortality. The package contains two datasets: `vulnerability` and `diabetes`, and relevant functions to conduct basic data cleaning and visualizations. 

## Example

This is a basic example which shows you how to solve a common problem:

- Attach required packages

```{r example}
library(svi)
library(dplyr)
library(purrr)
```

- Plot US SVI Map

```{r fig.width=9,fig.height=4}
vulnerability |> mnnn_to_na(names(which(map_lgl(vulnerability, is.double))), -999) |>
  rename(fips = FIPS) |>
  svi_map("RPL_THEMES")
```

- Plot US Diabetes Mortality Map

```{r fig.width=9,fig.height=4}
diabetes |> cr_interpolate(reliable = FALSE) |>
  rename(fips = County.Code) |>
  mortality_map("Crude.Rate")
```

- Scatter Plot of Diabetes Mortality vs. SVI Estimates

```{r fig.width=9,fig.height=4}
df <- prepare(vulnerability, diabetes, reliable = FALSE)
df |> mortality_vs_svi_scatter("RPL_THEMES", "Diabetes")
```

